---
title: "Project 2: Relationships between Mental Health, Physical Health, Inactivity, and Obesity in Counties in Texas"
author: "Taylor Gardiner tpg422"
output: html_document
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.align = "center", warning = F, message = F, R.options=list(max.print=100))

library(readr)
library(tidyverse)
library(plotROC)
```

### Introduction

##### For my project, I chose to analyze the way that mental health affects physical health throughout the state of Texas. I have always been interested in the multi-faceted effects of mental health and their lasting impact on overall health. I acquired my datasets through countyhealthrankings.org and limited my data to Texas because I wanted to keep my research local. I used four datasets that I believe created a fairly comprehensive look at mental and physical well-being. The first dataset I used contained information about the average number of poor mental health days in the last 30 days for adults over the age of 20, separated by county. The second dataset I used contained information about the average number of poor physical health days in the last 30 days for adults over the age of 20, separated by county. The third dataset I used showed the percentage of adults over 20 who reported no physical activity, separated by county. The fourth dataset I used showed the percentage of adults over 20 who report a BMI greater than or equal to 30 kg/m^2, also separated by county. Each dataset also contained data about the z-score and error margin associated with each county value. I expect to see that counties that report higher numbers of poor mental health days to also report higher numbers of poor physical health days, higher incidence of physical inactivity, and higher rates of obesity. I expect to see this because mental health can greatly impact an individual's diet and tendency to exercise. I had to tidy the data to join the columns together and create a new categorical variable. I also had to remove the percentage signs from the inactivity and obesity percentage data. This resulted in the dataset I used for Project 1 and will be using for Project 2. My project 1 dataset got deleted, so I recreated it below.

```{r}
# Recreating my Project 1 dataset
# Importing raw data
mental_health_days <- read_csv("~/Desktop/school stuff/comp bio/projects/project 1/csv files/mental_health_days.csv")

inactivity <- read_csv("~/Desktop/school stuff/comp bio/projects/project 1/csv files/inactivity.csv")

obesity <- read_csv("~/Desktop/school stuff/comp bio/projects/project 1/csv files/obesity.csv")

phys_health_days <- read_csv("~/Desktop/school stuff/comp bio/projects/project 1/csv files/phys_health_days.csv")

# Getting a glimpse at my datasets using the head() function
head(mental_health_days)
head(phys_health_days)
head(obesity)
head(inactivity)

# Tidying my data

# Tidying my poor mental health days data and saving it as a new dataset.
mentaldata <- mental_health_days %>% 
  rename(pmhd=`County Value `) %>% # Renaming the 'County Value' variable to 'pmhd' to represent poor mental health days
  select(County, pmhd) # Selecting only the county and pmhd columns for later use
head(mentaldata)

#Tidying my poor physical health days data and saving it as a new dataset.
physicaldata <- phys_health_days %>% 
  rename(pphd=`County Value `) %>% # Renaming the 'County Value' variable to 'pphd' to represent poor physical health days
  select(County, pphd) # Selecting only the county and pphd columns for later use
head(physicaldata)

#Tidying my obesity percentage data and saving it as a new dataset.
obesitydata <- obesity %>% 
  separate('County Value', into = c("obesity_percent", "percentages"), sep = 2) %>% # Removing the percentage signs from the 'County Value' column and renaming the column 'obesity_percent'
  select(County, obesity_percent) # Selecting only the county and obesity_percent columns for later use
head(obesitydata)

#Tidying my physical inactivity data and saving it as a new dataset.
inactivitydata <- inactivity %>% 
  separate('County Value', into = c("inactivity_percent", "percentages"), sep = 2) %>% # Removing the percentage signs from the 'County Value' column and renaming the column 'inactivity_percent'
  select(County, inactivity_percent) # Selecting only the county and inactivity_percent columns for later use
head(inactivitydata)

# Joining my datasets together
projectdataset <- mentaldata %>% 
  left_join(physicaldata, by="County") %>% # Adding poor physical health days data, matching by county name
  left_join(obesitydata, by="County") %>% # Adding obesity percentages data, matching by county name
  left_join(inactivitydata, by="County") # Adding physical inactivty percentages data, matching by county name

# Creating a new categorical variable for Obesity Level
fulldataset <- projectdataset %>%
   mutate(obesity_level=case_when(obesity_percent>=40 ~ "high", 
                                   30<=obesity_percent & obesity_percent<40 ~ "med",
                                   obesity_percent<30 ~ "low"))
head(fulldataset)

# To coerce my obesity_percent and inactivity_percent variables to numeric
fulldataset <- fulldataset %>% mutate_at(vars(obesity_percent, inactivity_percent), cols = NULL, as.numeric)
```


### Exploratory Data Analysis

#### First, I calculated some summary statistics to get a better sense of my dataset. 
```{r}
# To find mean and sd for my numeric variables at each level of my categorical variable
fulldataset %>% 
  group_by(obesity_level) %>% # To group by obesity levels
  summarize(mean_pmhd=mean(pmhd), sd_pmhd=sd(pmhd), mean_pphd=mean(pphd), sd_pphd=sd(pphd), mean_obesity=mean(obesity_percent), sd_obesity=sd(obesity_percent), mean_inactivity=mean(inactivity_percent), sd_inactivity=sd(inactivity_percent))
```
Obesity Level | Mean PMHD (days) | SD PMHD (days) | Mean PPHD (days) | SD PPHD (days) | Mean Obesity% | SD Obesity % | Mean Inactivity% | SD Inactivity %
------------- | --------- | ------- | --------- | ------- | ------------- | ------------ | ---------------- | -------------
High | 4.2 | .248 | 4.08 | .341 | 41.6 | 2.11 | 31.2 | 4.48
Medium | 4.1 | .246 | 3.99 | .45 | 33.8 | 2.74 | 28.7 | 4.43
Low | 3.99 | .282 | 3.87 | .45 | 26.4 | 2.01 | 25.1 | 3.91

*A table showing the mean and standard deviation of each numeric variable at each obesity level.*

```{r}
# To determine how many counties are in each obesity level category.
fulldataset %>% filter(obesity_level=="high") %>% summarize(count=n())
fulldataset %>% filter(obesity_level=="med") %>% summarize(count=n())
fulldataset %>% filter(obesity_level=="low") %>% summarize(count=n())
```
*20 counties had obesity percentages in the high category, 127 counties had obesity percentages in the medium category, and 107 counties had obesity percentages in the low category.*

#### To visualize my data, I created a few key graphs to identify potential significant relationships between my variables.
```{r}
# To create a correlation matrix
cormatrix <- fulldataset %>% select_if(is.numeric)

# To create a correlation heatmap 
cor(cormatrix, use="pairwise.complete.obs") %>% # To calculate correlation values
  as.data.frame %>% # Save as a data frame
  rownames_to_column %>% # Convert rownames to a specific variable
  pivot_longer(-1, names_to="other_var", values_to="correlation") %>% # Pivot so that all correlations are in the same column
  ggplot(aes(rowname, other_var, fill=correlation)) + geom_tile() + scale_fill_gradient2(low="red",mid="white",high="blue") + geom_text(aes(label=round(correlation,2)), color="black", size=4) + labs(title= "Correlation Matrix for Numeric Variables", x="Variable 1", y="Variable 2") + theme(axis.text.x = element_text(angle=45, hjust=1))
```
*It is clear from the correlation heatmap that the strongest relationship exists between the number of poor mental health days and the number of poor physical health days. I expected this because people who are not feeling well physically are most likely not feeling well mentally either, and vice versa. The weakest relationship exists between the number of poor physical health days and the percentage of people who did not engage in physical activity. I found this surprising because I assumed that people who were not feeling well physically would not want to engage in physical activity. There was a moderately strong relationship between the percentage of adults who were obese and the percentage of adults who did not engage in physical activity. This was expected because exercise is a way to combat obesity.*
```{r}
# Creating a scatterplot to visualize poor mental health and inactivity for different obesity levels
ggplot(data = fulldataset, aes(x = pmhd, y = inactivity_percent, color = obesity_level)) + geom_point(size=2) + theme_minimal() + ggtitle("Relationship between Poor Mental Health Days and Inactivity") + xlab("Poor Mental Health Days (days)") + ylab("Inactivity Percentage (%)") + geom_smooth(method="lm") + scale_color_brewer(palette="Spectral")
```
*It can be seen from the scatterplot that counties in the high obesity level category had relatively higher numbers of poor mental health days and a higher inactivity percentage. Counties in the low and medium obesity categories had more varied data, but it is clear that counties in the low obesity category had lower rates of inactivity. Interestingly, counties in the low and medium obesity categories seem to have similar incidence of poor mental health days. It can also be seen that the linear relationship between number of poor mental health days and inactivity percentage is stronger for the medium obesity level group. This is a relationship I plan to explore later through regression.*
```{r}
# Creating a bar graph to show poor mental health days per obesity level with error bars
ggplot(data=fulldataset, aes(x=obesity_level, fill=obesity_level)) + geom_bar(aes(y=pmhd), stat="summary", fun="mean") + scale_y_continuous(name="Number of Poor Mental Health Days per Month", breaks = seq(0,6, by=1)) + geom_errorbar(stat="summary", fun.data="mean_se", aes(y=pmhd), width=.5) + scale_fill_brewer(palette="RdPu")

# Creating a bar graph to show poor physical health days per obesity level with error bars
ggplot(data=fulldataset, aes(x=obesity_level, fill=obesity_level)) + geom_bar(aes(y=pphd), stat="summary", fun="mean") + scale_y_continuous(name="Number of Poor Physical Health Days per Month", breaks = seq(0,6, by=1)) + geom_errorbar(stat="summary", fun.data="mean_se", aes(y=pphd), width=.5) + scale_fill_brewer(palette="YlOrRd")
```
*It can be seen from these graphs that poor mental health days and poor physical health days follow similar trends through each obesity level. Counties with high obesity levels reported higher numbers of poor mental and physical health days and counties with low obesity levels reported lower numbers of poor mental health days. The graphs are also similar in that counties with high obesity levels have larger standard errors in their numbers of poor mental and physical health days. This fits my original expectation that poor mental health plays a role in poor physical health and obesity in Texas.*

### MANOVA

#### I wanted to conduct a MANOVA test analyzing differences in mean poor mental health days, poor physical health days, and inactivity percentages across the categories of my obesity level variable. I did not test obesity percentage values across the obesity level categories because there would clearly be a difference in mean obesity across the three levels.
```{r}
# Running a MANOVA test
manova_project2 <- manova(cbind(pmhd,pphd,inactivity_percent) ~ obesity_level, data = fulldataset)

# Viewing output of MANOVA test
summary(manova_project2)
```
*A one-way MANOVA was conducted to determine the effect of the obesity level (low, medium, high) on three dependent variables (Poor Mental Health Days, Poor Physical Health Days, Inactivity Percentage). Significant differences were found among the three obesity levels for at least one of the dependent variables (Pillai’s trace = 0.211, pseudo F(6,500) = 9.8743, p < .0001).*

#### Because I had a significant MANOVA result, I conducted univariate ANOVA tests and post-hoc t testing to determine which groups differ. 
```{r}
# One-way ANOVA for each variable
summary.aov(manova_project2)
```
*Univariate ANOVAs for each dependent variable were conducted as follow-up tests to the MANOVA. The one-way ANOVA test was also significant for poor mental health days (F(2,251) = 7.3671, p=.0008) and inactivity percentage (F(2,251) = 30.262, p < .0001). Poor physical health days no longer significantly differed among the three obesity level groups.*
```{r}
# Post-hoc t testing
# For poor mental health days
pairwise.t.test(fulldataset$pmhd,fulldataset$obesity_level, p.adj="none")

# For inactivity percentage
pairwise.t.test(fulldataset$inactivity_percent,fulldataset$obesity_level, p.adj="none")

# Calculating probability of Type I error
1-(.95^10)

# Bonferroni adjusted alpha
.05/10
```
*Throughout my MANOVA, ANOVA, and post-hoc t testing, I ran 10 tests. This resulted in a .401 probability of at least one Type I error. To fix this, I used a Bonferroni adjusted alpha level of .005. Post hoc analysis was performed conducting pairwise comparisons to determine which obesity levels differed in poor mental health days and inactivity percentage. After adjusting for multiple comparisons (Bonferroni alpha=.005), it was concluded that the low obesity category significantly differs from both the medium and high obesity categories in poor mental health days and inactivity percent, but that the medium and high obesity categories do not significantly differ from each other in poor mental health days and inactivity percent.* 

### Randomization Test
#### For my randomization test, I wanted to determine if there was a significant difference in poor mental health days for the medium and high obesity level groups. For this test, my null hypothesis is that there is no mean difference in poor mental health days between the two groups. My alternate hypothesis is that there is a significant difference in poor mental health days between the two groups.
```{r}
# Finding the true difference in poor mental health days between medium and high obesity levels
true_diff <- fulldataset %>% 
  filter(obesity_level=="med" | obesity_level=="high") %>% 
  group_by(obesity_level) %>% 
  summarize(means=mean(pmhd)) %>% 
  summarize(mean_diff=diff(means)) %>% 
  pull
true_diff

# Setting a seed for the randomization
set.seed(348)

# Create an empty vector to store the mean differences 
mean_diff <- vector()

# Create many randomizations with a for loop
for(i in 1:5000){ 
  temp <- data.frame(obesity = fulldataset$obesity_level, pmhd = sample(fulldataset$pmhd)) 
  
  mean_diff[i] <- temp %>% 
    group_by(obesity) %>%
    summarize(means = mean(pmhd)) %>%
    summarize(mean_diff = diff(means)) %>%
    pull
}

# Represent the distribution of the mean differences with a vertical line showing the true difference
{hist(mean_diff, main="Distribution of the mean differences"); abline(v = -.0997, col="red")}

# Calculating the corresponding two-sided p-value
mean(mean_diff > -true_diff | mean_diff < true_diff)
```
*Based on these results, I fail to reject the null hypothesis and do not find evidence to support that there is a mean difference in poor mental health days between the medium and high obesity level groups.*

### Linear Regression Model

#### For my linear regression model, I wanted to build a model predicting inactivity percent from obesity level and poor mental health days. 
```{r}
# Mean centering my numeric variables
fulldataset$pmhd_c <- fulldataset$pmhd - mean(fulldataset$pmhd)
head(fulldataset)

# Creating a graph to visualize the relationship between my variables
ggplot(data = fulldataset, aes(x = pmhd, y = inactivity_percent, color = obesity_level)) + geom_point(size=2) + theme_minimal() + ggtitle("Relationship between Poor Mental Health Days and Inactivity") + xlab("Poor Mental Health Days (days)") + ylab("Inactivity Percentage (%)") + geom_smooth(method="lm") + scale_color_brewer(palette="Spectral")
```
*This graph was originally created during my Exploratory Data Analysis, and the graph led me to my decision to explore these relationships through a linear model. It can be seen from the graph that for differing obesity levels, the correlation between poor mental health days and inactivity percent differs. This is hopefully a significant interaction. It can also be seen from this graph that poor mental health days and inactivity are positively correlated, so I hope that this is a significant effect as well.*

```{r}
# Creating my linear regression model
fit1 <- lm(inactivity_percent ~ pmhd_c * obesity_level, data = fulldataset)
summary(fit1)
```
*Texas counties with a high obesity level and mean poor mental health days have a predicted inactivity percent of 30.94%. For counties with a high obesity level, for every 1 unit increase in poor mental health days, inactivity percent goes up by 1.552% on average. Predicted inactivity percent for low obesity counties with average poor mental health days is 5.689% lower than high obesity counties with average poor mental health. Predicted inactivity percent for medium obesity counties with average poor mental health is 2.367% lower than high obesity counties with average poor mental health. The slope of poor mental health days on inactivity percent is 1.075 greater for low obesity counties than high obesity counties. The slope of poor mental health days on inactivity percent is 2.506 greater for medium obesity counties than high obesity counties. 22.78% of the variation in inactivity percent can be accounted for by poor mental health days, obesity level, and their interaction.* 
```{r}
# Checking linearity assumption
plot(fulldataset$pmhd_c,fulldataset$inactivity_percent)

# Checking residuals
plot(fit1, which = 1)

# Checking QQ plot for residuals
plot(fit1, which = 2)

# Checking normality with Shapiro-Wilk Test
shapiro.test(fit1$residuals)

# Checking homoscedasticity with Breusch-Pagan Test
library(sandwich)
library(lmtest)
bptest(fit1)
```
*Graphically, the plot and residuals seem mostly linear and pass the linearity assumption. The QQ plot of residuals appears normal except for a couple of outliers. After conducting the Shapiro-Wilk test, I conclude that the data is not normal (p-value=.03). After conducting the Breusch-Pagan test, I can affirm the homoscedasticity asssumption (p-value=.54).*
```{r}
# Robust Standard Errors
coeftest(fit1, vcov = vcovHC(fit1))
```
*Before computing robust SEs, the intercept and obesity level effects were significant, meaning that counties with a low or medium obesity level and average poor mental health days had a significantly lower inactivity percentage than counties with a high obesity level and average poor mental health days (t=-4.977, df=248, p-value<.001; t=-2.097, df=248, p-value=.037). After computing robust SEs, the effect of medium obesity level was no longer significant. Counties with a low obesity level and average poor mental health days had a significantly lower inactivity percentage than counties with a high obesity level and average poor mental health days (t=-4.4628, p-value<.001).*
```{r}
# Bootstapped Standard Errors
samp_SEs <- replicate(5000, {
  # Bootstrap the data
  boot_data <- sample_frac(fulldataset, replace = TRUE)
  # Fit regression model
  fitboot <- lm(inactivity_percent ~ pmhd_c * obesity_level, data = boot_data)
  # Save the coefficients
  coef(fitboot)
})

# Estimated SEs
samp_SEs %>%
  # Transpose the obtained matrices
  t %>%
  # Consider the matrix as a data frame
  as.data.frame %>%
  # Compute the standard error (standard deviation of the sampling distribution)
  summarize_all(sd)
```
*My bootstrapped standard errors did not differ much from my original and robust standard errors. The largest differences in standard errrors were for my mean centered poor mental health days and the two interaction terms. The robust and bootstrapped standard errors were similar for these terms, but both were different from the standard errors from the original linear regression.*

### Logistic Regression
#### For my logistic regression, I first created a new binary categorical variable that classified counties as either "inactive" or "active". Then I created a regression model that predicted this new variable based on poor mental health days and poor physical health days.
```{r}
# Creating a new binary categorical variable for inactivity level
fulldataset <- fulldataset %>%
    mutate(inactivity_level=case_when(inactivity_percent>=30 ~ "inactive",
                                    inactivity_percent<30 ~ "active"))

# Coding my binary variable
fulldataset <- fulldataset %>%
  mutate(y = ifelse(inactivity_level == "inactive", 1, 0))

# Logistic Regression Model
fit2 <- glm(y ~ pmhd + pphd,  data = fulldataset, family = "binomial")
summary(fit2)
exp(coef(fit2))
```
*Every one unit increase in poor mental health days increases the log odds of inactivity by 4.0943. Every one unit increase in poor mental health days multiplies the odds of inactivity by 59.99. Every one unit increase in poor physical health days decreases the log odds of inactivity by 1.8417. Every one unit increase in poor physical health days multiplies the odds of inactivity by .0159.*

```{r}
# Adding predictions to dataset
fulldataset$prob <- predict(fit2, type = "response")
fulldataset$predicted <- ifelse(fulldataset$prob > .5, "inactive", "active")
# Creating a confusion matrix
table(truth = fulldataset$inactivity_level, prediction = fulldataset$predicted)
# Computing Accuracy (correctly classified cases)
(166+5)/254
# Computing Sensitivity (true positive rate)
5/79
# Computing Specificity (true negative rate)
166/175
# Computing Precision (positive predicted value)
5/14
```
*My model correctly predicted counties' inactivity level 67% of the time. It had a True Positive Rate of 6.3% and a True Negative Rate of 94.86%. Its positive predicted value was 35.71%, meaning that of all the counties it predicted to have a positive result (inactive), it was correct 35.71% of those instances.*

```{r}
# Saving the predicted log-odds in the dataset
fulldataset$logit <- predict(fit2)

# Comparing to the outcome in dataset with a density plot
ggplot(fulldataset, aes(logit, fill = as.factor(inactivity_level))) +
  geom_density(alpha = .3) +
  geom_vline(xintercept = 0, lty = 2) +
  labs(fill = "Inactivity Classification")
```
*The darker portion of the graph represents counties that have been misclassified. Their predicted inactivity level was not accurate. It is clear from my graph that a fair amount of counties' inactivity level was inaccurately predicted. My model is not entirely accurate.*

#### Next, I created an ROC plot and calculated the AUC.
```{r}
# Creating the ROC plot
library(plotROC)
ROCplot <- ggplot(fulldataset) + 
   geom_roc(aes(d = y, m = prob), cutoffs.at = list(0.1, 0.5, 0.9))
ROCplot

# Calculating the AUC
calc_auc(ROCplot)
```
*My ROC plot and AUC show that my logistic regression model is a poor predictor of inactivity level. A randomly selected county from the inactive group has a test value larger than a randomly selected county from the active group 65.64% of the time.* 
